---
title: "The 'elo' Package"
author: "Ethan Heinzen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{The 'elo' Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# The `elo` Package

The `elo` package includes functions to address all kinds of Elo calculations.

```{r}
library(elo)
```

## Naming Schema

Most functions begin with the prefix "elo.", for easy autocompletion.

- Vectors or scalars of Elo scores are denoted `elo.A` or `elo.B`.

- Vectors or scalars of wins by team A are denoted by `wins.A`.

- Vectors or scalars of win probabilities are denoted by `p.A`.

- Vectors of team names are denoted `team.A` or `team.B`.

# Basic Functions

To calculate the probability team.A beats team.B, use `elo.prob()`:

```{r}
elo.A <- c(1500, 1500)
elo.B <- c(1500, 1600)
elo.prob(elo.A, elo.B)
```

To calculate the score update after the two teams play, use `elo.update()`:

```{r}
wins.A <- c(1, 0)
elo.update(wins.A, elo.A, elo.B, k = 20)
```

To calculate the new Elo scores after the update, use `elo.calc()`:

```{r}
elo.calc(wins.A, elo.A, elo.B, k = 20)
```

# The `elo.run()` function

## With two variable Elos

To calculate a series of Elo updates, use `elo.run()`. This function has a `formula = ` and 
`data = ` interface. We first load the dataset `tournament`.

```{r}
data(tournament)
str(tournament)
```

`formula = ` should be in the format of `wins.A ~ team.A + team.B`. The `score()` function
will help to calculate winners on the fly (1 = win, 0.5 = tie, 0 = loss).

```{r}
tournament$wins.A <- tournament$points.Home > tournament$points.Visitor
elo.run(wins.A ~ team.Home + team.Visitor, data = tournament, k = 20)
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament, k = 20)
```

For more complicated Elo updates, you can include the special function `k()` in the
`formula = ` argument. Here we're taking the log of the win margin as part of our update.

```{r}
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor +
        k(20*log(abs(points.Home - points.Visitor) + 1)), data = tournament)
```

It's also possible to adjust one team's Elo for a variety of factors
(e.g., home-field advantage). The `adjust()` special function will take as its second
argument a vector or a constant.

```{r}
elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
        data = tournament, k = 20)
```

## With a fixed-Elo opponent

`elo.run()` also recognizes if the second column is numeric,
and interprets that as a fixed-Elo opponent.

```{r}
tournament$elo.Visitor <- 1500
elo.run(score(points.Home, points.Visitor) ~ team.Home + elo.Visitor,
        data = tournament, k = 20)
```

## Regress Elos back to the mean

The special function `regress()` can be used to regress Elos back to a fixed value
after certain matches. Giving a logical vector identifies these matches after which to
regress back to the mean. Giving any other kind of vector regresses after the appropriate
groupings (see, e.g., `duplicated(..., fromLast = TRUE)`). The other two arguments determine
what Elo to regress to (`to = `), and by how much to regress toward that value (`by = `).

```{r}
tournament$elo.Visitor <- 1500
elo.run(score(points.Home, points.Visitor) ~ team.Home + elo.Visitor +
        regress(half, 1500, 0.2),
        data = tournament, k = 20)
```

## Group matches

The special function `group()` doesn't affect `elo.run()`, but determines matches
to group together in `as.matrix()` (below).

## Helper functions

There are several helper functions that are useful to use when interacting with
objects of class `"elo.run"`.

`summary.elo.run()` reports some summary statistics.

```{r}
e <- elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor,
             data = tournament, k = 20)
summary(e)
```

`as.matrix.elo.run()` creates a matrix of running Elos.

```{r}
head(as.matrix(e))
```

`as.data.frame.elo.run()` gives the long version (perfect, for, e.g., `ggplot2`).

```{r}
str(as.data.frame(e))
```

Finally, `final.elos()` will extract the final Elos per team.

```{r}
final.elos(e)
```

## Making Predictions

It is also possible to use the Elos calculated by `elo.run()` to make predictions on future match-ups.

```{r}
results <- elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
                   data = tournament, k = 20)
newdat <- data.frame(
  team.Home = "Athletic Armadillos",
  team.Visitor = "Blundering Baboons"
)
predict(results, newdata = newdat)
```

# Basic Functions Revisited - Formula Interface

All three of the "basic" functions accept formulas as input, just like
`elo.run()`.

```{r}
dat <- data.frame(elo.A = c(1500, 1500), elo.B = c(1500, 1600),
                  wins.A = c(1, 0), k = 20)
form <- wins.A ~ elo.A + elo.B + k(k)
elo.prob(form, data = dat)
elo.update(form, data = dat)
elo.calc(form, data = dat)
```

Note that for `elo.prob()`, `formula = ` can be more succinct:

```{r}
elo.prob(~ elo.A + elo.B, data = dat)
```

We can even adjust the Elos:

```{r}
elo.calc(wins.A ~ adjust(elo.A, 10) + elo.B + k(k), data = dat)
```
