---
title: "The 'elo' Package"
author: "Ethan Heinzen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{The 'elo' Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# The `elo` Package

The `elo` package includes functions to address all kinds of Elo calculations.

```{r}
library(elo)
```

## Naming Schema

Most functions begin with the prefix "elo.", for easy autocompletion.

- Vectors or scalars of Elo scores are denoted `elo.A` or `elo.B`.

- Vectors or scalars of wins by team A are denoted by `wins.A`.

- Vectors or scalars of win probabilities are denoted by `p.A`.

- Vectors of team names are denoted `team.A` or `team.B`.

# Basic Functions

To calculate the probability team.A beats team.B, use `elo.prob()`:

```{r}
elo.A <- c(1500, 1500)
elo.B <- c(1500, 1600)
elo.prob(elo.A, elo.B)
```

To calculate the score update after the two teams play, use `elo.update()`:

```{r}
wins.A <- c(1, 0)
elo.update(wins.A, elo.A, elo.B, k = 20)
```

To calculate the new Elo scores after the update, use `elo.calc()`:

```{r}
elo.calc(wins.A, elo.A, elo.B, k = 20)
```

# The `elo.run()` function

## With two variable Elos

To calculate a series of Elo updates, use `elo.run()`. This function has a `formula = ` and 
`data = ` interface. We first load the dataset `tournament`.

```{r}
data(tournament)
str(tournament)
```

`formula = ` should be in the format of `wins.A ~ team.A + team.B`. The `score()` function
will help to calculate winners on the fly (1 = win, 0.5 = tie, 0 = loss).

```{r}
tournament$wins.A <- tournament$points.Home > tournament$points.Visitor
elo.run(wins.A ~ team.Home + team.Visitor, data = tournament, k = 20)
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament, k = 20)
```

For more complicated Elo updates, you can include the special function `k()` in the
`formula = ` argument. Here we're taking the log of the win margin as part of our update.

```{r}
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor +
        k(20*log(abs(points.Home - points.Visitor) + 1)), data = tournament)
```

It's also possible to adjust one team's Elo for a variety of factors
(e.g., home-field advantage). The `adjust()` special function will take as its second
argument a vector or a constant.

```{r}
elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
        data = tournament, k = 20)
```

## With a fixed-Elo opponent

`elo.run()` also recognizes if the second column is numeric,
and interprets that as a fixed-Elo opponent.

```{r}
tournament$elo.Visitor <- 1500
elo.run(score(points.Home, points.Visitor) ~ team.Home + elo.Visitor,
        data = tournament, k = 20)
```

## Regress Elos back to the mean

The special function `regress()` can be used to regress Elos back to a fixed value
after certain matches. Giving a logical vector identifies these matches after which to
regress back to the mean. Giving any other kind of vector regresses after the appropriate
groupings (see, e.g., `duplicated(..., fromLast = TRUE)`). The other three arguments determine
what Elo to regress to (`to = `, which could be a different value for different teams),
by how much to regress toward that value (`by = `), and whether to regress teams which aren't
actively playing (`regress.unused = `).

```{r}
tournament$elo.Visitor <- 1500
elo.run(score(points.Home, points.Visitor) ~ team.Home + elo.Visitor +
        regress(half, 1500, 0.2),
        data = tournament, k = 20)
```

## Group matches

The special function `group()` doesn't affect `elo.run()`, but determines matches
to group together in `as.matrix()` (below).

## Helper functions

There are several helper functions that are useful to use when interacting with
objects of class `"elo.run"`.

`summary.elo.run()` reports some summary statistics.

```{r}
e <- elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor,
             data = tournament, k = 20)
summary(e)
```

`as.matrix.elo.run()` creates a matrix of running Elos.

```{r}
head(as.matrix(e))
```

`as.data.frame.elo.run()` gives the long version (perfect, for, e.g., `ggplot2`).

```{r}
str(as.data.frame(e))
```

Finally, `final.elos()` will extract the final Elos per team.

```{r}
final.elos(e)
```

## Making Predictions

It is also possible to use the Elos calculated by `elo.run()` to make predictions on future match-ups.

```{r}
results <- elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
                   data = tournament, k = 20)
newdat <- data.frame(
  team.Home = "Athletic Armadillos",
  team.Visitor = "Blundering Baboons"
)
predict(results, newdata = newdat)
```

# Basic Functions Revisited - Formula Interface

All three of the "basic" functions accept formulas as input, just like
`elo.run()`.

```{r}
dat <- data.frame(elo.A = c(1500, 1500), elo.B = c(1500, 1600),
                  wins.A = c(1, 0), k = 20)
form <- wins.A ~ elo.A + elo.B + k(k)
elo.prob(form, data = dat)
elo.update(form, data = dat)
elo.calc(form, data = dat)
```

Note that for `elo.prob()`, `formula = ` can be more succinct:

```{r}
elo.prob(~ elo.A + elo.B, data = dat)
```

We can even adjust the Elos:

```{r}
elo.calc(wins.A ~ adjust(elo.A, 10) + elo.B + k(k), data = dat)
```

# Comparison Models

## Logistic Regression

It's also possible to compare teams' skills using logistic regression. A matrix of dummy variables is constructed, one for each team,
where a value of 1 indicates a home team and -1 indicates a visiting team. The intercept then indicates a home-field advantage.
Including an adjustment using `adjust()` in the formula also includes that in the model.

```{r}
results <- elo.glm(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                   subset = points.Home != points.Visitor) # to get rid of ties for now
summary(results)
predict(results, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE))

tournament$neutral <- replace(rep(0, nrow(tournament)), 30:35, 1)
summary(elo.glm(score(points.Home, points.Visitor) ~ adjust(team.Home, neutral) + team.Visitor + group(week),
                data = tournament, subset = points.Home != points.Visitor))
```

The models can be built "running", where predictions for the next group of games are made based on past data. Consider using the `skip=` argument
to skip the first few groups (otherwise the model might have trouble converging).

Note that predictions from this object use a model fit on all the data.

```{r}
results <- elo.glm(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                   subset = points.Home != points.Visitor, running = TRUE, skip = 5)
summary(results)
predict(results, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE)) # the same thing
```

## Markov Chain

It's also possible to compare teams' skills using a Markov-chain-based model, as outlined in
[Kvam and Sokol (2006)](https://scholarship.richmond.edu/cgi/viewcontent.cgi?article=1207&context=mathcs-faculty-publications).
In short, imagine a judge who randomly picks one of two teams in a matchup, where the winner gets chosen with
probability p (here, for convenience, 'k') and the loser with probability 1-p (1-k). This forms a transition
matrix, whose stationary distribution gives a ranking of teams. The differences in ranking are then fed
into a logistic regession model to predict win status. Any adjustments made using `adjust()` are also included
in this logistic regression. In this way, you could adjust the intercept for games played on neutral fields (for
example).

Note that by assigning probabilities in the right way, this function emits the Logistic Regression Markov Chain
model (LRMC). Consider using the in-formula function `k()` for this.

```{r}
mc <- elo.markovchain(score(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament,
                      subset = points.Home != points.Visitor, k = 0.7)
summary(mc)
predict(mc, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE))
summary(elo.markovchain(score(points.Home, points.Visitor) ~ adjust(team.Home, neutral) + team.Visitor,
                        data = tournament, subset = points.Home != points.Visitor, k = 0.7))
```

These models can also be built "running", where predictions for the next group of games are made based on past data.
Consider using the `skip=` argument to skip the first few groups (otherwise the model might have trouble converging).

Note that predictions from this object use a model fit on all the data.

```{r}
mc <- elo.markovchain(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                      subset = points.Home != points.Visitor, k = 0.7, running = TRUE, skip = 5)
summary(mc)
predict(mc, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE)) # the same thing
```
