---
title: "The 'elo' Package"
author: "Ethan Heinzen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{The 'elo' Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# The `elo` Package

The `elo` package includes functions to address all kinds of Elo calculations.

```{r}
library(elo)
```

## Naming Schema

Most functions begin with the prefix "elo.", for easy autocompletion.

- Vectors or scalars of Elo scores are denoted `elo.A` or `elo.B`.

- Vectors or scalars of wins by team A are denoted by `wins.A`.

- Vectors or scalars of win probabilities are denoted by `p.A`.

- Vectors of team names are denoted `team.A` or `team.B`.

# Basic Functions

To calculate the probability team.A beats team.B, use `elo.prob()`:

```{r}
elo.A <- c(1500, 1500)
elo.B <- c(1500, 1600)
elo.prob(elo.A, elo.B)
```

To calculate the score update after the two teams play, use `elo.update()`:

```{r}
wins.A <- c(1, 0)
elo.update(wins.A, elo.A, elo.B, k = 20)
```

To calculate the new Elo scores after the update, use `elo.calc()`:

```{r}
elo.calc(wins.A, elo.A, elo.B, k = 20)
```

# The `elo.run()` function

## With two variable Elos

To calculate a series of Elo updates, use `elo.run()`. This function has a `formula = ` and 
`data = ` interface. We first load the dataset `tournament`.

```{r}
data(tournament)
str(tournament)
```

`formula = ` should be in the format of `wins.A ~ team.A + team.B`. The `score()` function
will help to calculate winners on the fly (1 = win, 0.5 = tie, 0 = loss).

```{r}
tournament$wins.A <- tournament$points.Home > tournament$points.Visitor
elo.run(wins.A ~ team.Home + team.Visitor, data = tournament, k = 20)
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament, k = 20)
```

For more complicated Elo updates, you can include the special function `k()` in the
`formula = ` argument. Here we're taking the log of the win margin as part of our update.

```{r}
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor +
        k(20*log(abs(points.Home - points.Visitor) + 1)), data = tournament)
```

You can also adjust the home and visitor teams with different k's:

```{r}
k1 <- 20*log(abs(tournament$points.Home - tournament$points.Visitor) + 1)
elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + k(k1, k1/2), data = tournament)
```

It's also possible to adjust one team's Elo for a variety of factors
(e.g., home-field advantage). The `adjust()` special function will take as its second
argument a vector or a constant.

```{r}
elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
        data = tournament, k = 20)
```

## With a fixed-Elo opponent

`elo.run()` also recognizes if the second column is numeric,
and interprets that as a fixed-Elo opponent.

```{r}
tournament$elo.Visitor <- 1500
elo.run(score(points.Home, points.Visitor) ~ team.Home + elo.Visitor,
        data = tournament, k = 20)
```

## Regress Elos back to the mean

The special function `regress()` can be used to regress Elos back to a fixed value
after certain matches. Giving a logical vector identifies these matches after which to
regress back to the mean. Giving any other kind of vector regresses after the appropriate
groupings (see, e.g., `duplicated(..., fromLast = TRUE)`). The other three arguments determine
what Elo to regress to (`to = `, which could be a different value for different teams),
by how much to regress toward that value (`by = `), and whether to regress teams which aren't
actively playing (`regress.unused = `).

```{r}
tournament$elo.Visitor <- 1500
elo.run(score(points.Home, points.Visitor) ~ team.Home + elo.Visitor +
        regress(half, 1500, 0.2),
        data = tournament, k = 20)
```

## Group matches

The special function `group()` doesn't affect `elo.run()`, but determines matches
to group together in `as.matrix()` (below).

## Helper functions

There are several helper functions that are useful to use when interacting with
objects of class `"elo.run"`.

`summary.elo.run()` reports some summary statistics.

```{r}
e <- elo.run(score(points.Home, points.Visitor) ~ team.Home + team.Visitor,
             data = tournament, k = 20)
summary(e)
rank.teams(e)
```

`as.matrix.elo.run()` creates a matrix of running Elos.

```{r}
head(as.matrix(e))
```

`as.data.frame.elo.run()` gives the long version (perfect, for, e.g., `ggplot2`).

```{r}
str(as.data.frame(e))
```

Finally, `final.elos()` will extract the final Elos per team.

```{r}
final.elos(e)
```

## Making Predictions

It is also possible to use the Elos calculated by `elo.run()` to make predictions on future match-ups.

```{r}
results <- elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
                   data = tournament, k = 20)
newdat <- data.frame(
  team.Home = "Athletic Armadillos",
  team.Visitor = "Blundering Baboons"
)
predict(results, newdata = newdat)
```

## Advanced: custom probability and updates

We now get to `elo.run2()`, a copy of `elo.run()` (but implemented in R) that allows for custom probability calculations and Elo updates.

For instance, suppose you want to change the adjustment based on team A's current Elo:

```{r}
custom_update <- function(wins.A, elo.A, elo.B, k, adjust.A, adjust.B, ...)
{
  k*(wins.A - elo.prob(elo.A, elo.B, adjust.B = adjust.B,
                       adjust.A = ifelse(elo.A > 1500, adjust.A / 2, adjust.A)))
}
custom_prob <- function(elo.A, elo.B, adjust.A, adjust.B)
{
  1/(1 + 10^(((elo.B + adjust.B) - (elo.A + ifelse(elo.A > 1500, adjust.A / 2, adjust.A)))/400.0))
}
er2 <- elo.run2(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
                data = tournament, k = 20, prob.fun = custom_prob, update.fun = custom_update)
final.elos(er2)
```

Compare this to the results from the default:

```{r}
er3 <- elo.run(score(points.Home, points.Visitor) ~ adjust(team.Home, 10) + team.Visitor,
               data = tournament, k = 20)
final.elos(er3)
```

This example is a bit contrived, as it'd be easier just to use `adjust()` (actually, this is tested for in the tests), but the point remains.


# Basic Functions Revisited - Formula Interface

All three of the "basic" functions accept formulas as input, just like
`elo.run()`.

```{r}
dat <- data.frame(elo.A = c(1500, 1500), elo.B = c(1500, 1600),
                  wins.A = c(1, 0), k = 20)
form <- wins.A ~ elo.A + elo.B + k(k)
elo.prob(form, data = dat)
elo.update(form, data = dat)
elo.calc(form, data = dat)
```

Note that for `elo.prob()`, `formula = ` can be more succinct:

```{r}
elo.prob(~ elo.A + elo.B, data = dat)
```

We can even adjust the Elos:

```{r}
elo.calc(wins.A ~ adjust(elo.A, 10) + elo.B + k(k), data = dat)
```

# Comparison Models

## Win/Loss Logistic Regression

The first model computes teams' win percentages, and feeds the differences of percentages into a regression.
Including an adjustment using `adjust()` in the formula also includes that in the model.
You could also adjust the intercept for games played on neutral fields by using the `neutral()` function.

```{r}
e.winpct <- elo.winpct(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                       subset = points.Home != points.Visitor) # to get rid of ties for now
summary(e.winpct)
rank.teams(e.winpct)
predict(e.winpct, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE))

tournament$neutral <- replace(rep(0, nrow(tournament)), 30:35, 1)
summary(elo.winpct(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + neutral(neutral) + group(week),
                   data = tournament, subset = points.Home != points.Visitor))
```

The models can be built "running", where predictions for the next group of games are made based on past data. Consider using the `skip=` argument
to skip the first few groups (otherwise the model might have trouble converging).

Note that predictions from this object use a model fit on all the data.

```{r}
e.winpct <- elo.winpct(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                       subset = points.Home != points.Visitor, running = TRUE, skip = 5)
summary(e.winpct)
predict(e.winpct, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE)) # the same thing
```

## Logistic Regression

It's also possible to compare teams' skills using logistic regression. A matrix of dummy variables is constructed, one for each team,
where a value of 1 indicates a home team and -1 indicates a visiting team. The intercept then indicates a home-field advantage.
To denote games played in a neutral setting (that is, without home-field advantage), use the `neutral()` function.
In short, the intercept will then be set to `1 - neutral()`.
Including an adjustment using `adjust()` in the formula also includes that in the model.

```{r}
results <- elo.glm(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                   subset = points.Home != points.Visitor) # to get rid of ties for now
summary(results)
rank.teams(results)
predict(results, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE))

summary(elo.glm(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + neutral(neutral) + group(week),
                data = tournament, subset = points.Home != points.Visitor))
```

The models can be built "running", where predictions for the next group of games are made based on past data. Consider using the `skip=` argument
to skip the first few groups (otherwise the model might have trouble converging).

Note that predictions from this object use a model fit on all the data.

```{r}
results <- elo.glm(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                   subset = points.Home != points.Visitor, running = TRUE, skip = 5)
summary(results)
predict(results, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE)) # the same thing
```

## Markov Chain

It's also possible to compare teams' skills using a Markov-chain-based model, as outlined in
[Kvam and Sokol (2006)](https://scholarship.richmond.edu/cgi/viewcontent.cgi?article=1207&context=mathcs-faculty-publications).
In short, imagine a judge who randomly picks one of two teams in a matchup, where the winner gets chosen with
probability p (here, for convenience, 'k') and the loser with probability 1-p (1-k). In other words,
we assume that the probability that the winning team is better than the losing team given that it won is
k, and the probability that the losing team is better than the winning team given that it lost is (1-k).
This forms a transition matrix, whose stationary distribution gives a ranking of teams.
The differences in ranking are then fed into a logistic regession model to predict win status.
Any adjustments made using `adjust()` are also included in this logistic regression.
You could also adjust the intercept for games played on neutral fields by using the `neutral()` function.

```{r}
mc <- elo.markovchain(score(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament,
                      subset = points.Home != points.Visitor, k = 0.7)
summary(mc)
rank.teams(mc)
predict(mc, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE))
summary(elo.markovchain(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + neutral(neutral),
                        data = tournament, subset = points.Home != points.Visitor, k = 0.7))
```

These models can also be built "running", where predictions for the next group of games are made based on past data.
Consider using the `skip=` argument to skip the first few groups (otherwise the model might have trouble converging).

Note that predictions from this object use a model fit on all the data.

```{r}
mc <- elo.markovchain(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                      subset = points.Home != points.Visitor, k = 0.7, running = TRUE, skip = 5)
summary(mc)
predict(mc, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE)) # the same thing
```

### A note about LRMC

Note that by assigning probabilities in the right way, this function emits the Logistic Regression Markov Chain
model (LRMC). Use the in-formula function `k()` for this. IMPORTANT: note that `k()` denotes the probability
assigned to the *winning* team, not the *home* team (for instance). If `rH(x)` denotes the probability that
the home team is better given that they scored `x` points more than the visiting team (allowing for x to be negative),
then an LRMC model might look something like this:

```{r eval=FALSE}
elo.markovchain(floor(wins.home) ~ team.home + team.visitor + k(ifelse(x > 0, rH(x), 1 - rH(x))))
```

Why do we use `floor()` here? This takes care of the odd case where teams tie. In this case, `rH(x) < 0.5` because
we expected the home team to win by virtue of being home. By default, `elo.markovchain()` will split any ties down
the middle (i.e., 0.5 and 0.5 instead of p and 1-p), which isn't what we want; we want the visiting team to get
a larger share than the home team. Telling `elo.markovchain()` that the visiting team "won" gives the visiting team
its whole share of `p`.

Alternatively, if `h` denotes a home-field advantage (in terms of score), the model becomes:

```{r eval=FALSE}
elo.markovchain(ifelse(home.points - visitor.points > h, 1, 0) ~ team.home + team.visitor + k(pmax(rH(x), 1 - rH(x))))
```

In this case, the home team "won" if it scored more than `h` points more than the visiting team. Since
`rH(x) > 0.5` if `x > h`, then `pmax()` will assign the proper probability to the pseudo-winning team.

Finally, do note that using `neutral()` isn't sufficient for adjusting for games played on neutral ground,
because the adjustment is only taken into account in the logistic regression to produce probabilities, not
the building of the transition matrix. Therefore, you'll want to also account for neutral wins/losses in
`k()` as well.

## Colley Matrix Method

It's also possible to compare teams' skills using the Colley Matrix method, as outlined in
[Colley (2002)](http://www.colleyrankings.com/matrate.pdf).
The coefficients to the Colley matrix formulation gives a ranking of teams.
The differences in ranking are then fed into a logistic regession model to predict win status.
Here 'k' denotes how convincing a win is; it represents the fraction of the win assigned to the winning
team and the fraction of the loss assigned to the losing team. Setting 'k' = 1 emits the bias-free method
presented by Colley.
Any adjustments made using `adjust()` are also included in this logistic regression.
You could also adjust the intercept for games played on neutral fields by using the `neutral()` function.

```{r}
co <- elo.colley(score(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament,
                 subset = points.Home != points.Visitor)
summary(co)
rank.teams(co)
predict(co, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE))
summary(elo.colley(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + neutral(neutral),
                   data = tournament, subset = points.Home != points.Visitor))
```

These models can also be built "running", where predictions for the next group of games are made based on past data.
Consider using the `skip=` argument to skip the first few groups (otherwise the model might have trouble converging).

Note that predictions from this object use a model fit on all the data.

```{r}
co <- elo.colley(score(points.Home, points.Visitor) ~ team.Home + team.Visitor + group(week), data = tournament,
                      subset = points.Home != points.Visitor, running = TRUE, skip = 5)
summary(co)
predict(co, newdata = data.frame(team.Home = "Athletic Armadillos", team.Visitor = "Blundering Baboons", stringsAsFactors = FALSE)) # the same thing
```

# Modeling Margin of Victory Instead of Wins

`elo.glm()`, `elo.markovchain()`, and `elo.winpct()` all allow for modeling of margins of
victory instead of simple win/loss using the `mov()` function. Note that one must set the
`family="gaussian"` argument to get linear regression instead of logistic regression.

```{r}
summary(elo.glm(mov(points.Home, points.Visitor) ~ team.Home + team.Visitor, data = tournament,
                family = "gaussian"))
```
